---
title: EDS 222 HW1
subtitle: Foundations of inference
author: Courtney Lorey
---

## 1. Conduct a permutation test

Conduct a permutation test to investigate the following question:

> Do homes built pre-1970 have higher average lead soil concentrations?

To perform this test, you'll need the `exide-property-covariates.csv` file found in `/courses/EDS222/data`. This file contains the year each property was built.

::: {.callout-note}

The question is about _average lead soil concentrations_. You'll need to first get the average lead soil concentration for each property, then join your result to the property covariate data to get the year the house was built.

:::

$H_0$: There is no difference in average lead soil concentrations between homes built before and after 1970.

$H_A$: Homes built pre-1970 have higher average lead soil concentrations.

```{r}
# Load packages, set seed
library(tidyverse)
set.seed(222)

# Load data
properties <- read_csv('/courses/EDS222/data/exide-property-covariates.csv')
soil <- read_csv('/courses/EDS222/data/exide-soil-cores.csv')

# Remove properties that don't have a year_built
properties <- properties |> 
    drop_na(year_built) # This removed 10 NA rows

# Get the average lead soil concentration for each property
avg_ppm <- soil |> 
    summarize(mean_pb_ppm = mean(pb_ppm), .by = property_id)

# Join data
soil_cores <- left_join(properties, avg_ppm, by = join_by(property_id))

# Add a column that distinguishes pre and post 1970
soil_cores <- soil_cores |>
    mutate(pre1970 = (year_built < 1970)) 

# Summarize to get the difference in mean lead ppm between pre and post 1970 houses
soil_means_pre_post <- soil_cores |> 
    summarize(mean_pre_post = mean(mean_pb_ppm), .by = pre1970) |> 
    arrange(pre1970)

# Find the observed difference in means
obs_diff_means <- soil_means_pre_post$mean_pre_post[2] - soil_means_pre_post$mean_pre_post[1]

# Shuffle the predictor variable and recalculate the statistic 
shuffled_year <- soil_cores |>
  mutate(pre1970 = sample(pre1970)) |>
  summarize(mean_pre_post = mean(mean_pb_ppm), .by = pre1970) |> 
  arrange(pre1970)

shuffled_diff_means <- shuffled_year$mean_pre_post[2] -
  shuffled_year$mean_pre_post[1]

# Repeat shuffle
year_null_dist <- replicate(
  1e4,
  {
    shuffled_year <- soil_cores |>
        mutate(pre1970 = sample(pre1970)) |>
        summarize(mean_pre_post = mean(mean_pb_ppm), .by = pre1970) |> 
        arrange(pre1970)
        
    shuffled_year$mean_pre_post[2] - shuffled_year$mean_pre_post[1]
  }
)

# Visualize the null vs. the observed value
tibble(diff_means = year_null_dist) |>
  ggplot(aes(diff_means)) +
  geom_histogram(binwidth = 5) + 
  geom_vline(xintercept = obs_diff_means, color = 'magenta')

# Calculate the p-value
pval <- mean(year_null_dist >= obs_diff_means)
pval

```

### Interpreting our results

Based on our p-value of 0, which is less than the significance level (α = .05), we reject the null hypothesis that there is no difference in average lead soil concentrations between homes built before and after 1970, and have sufficient evidence to support the alternate hypothesis, that homes built pre-1970 have higher average lead soil concentrations.

## 2. Estimate a confidence interval

Estimate a confidence interval to investigate the following question:

> How much is the difference in average lead soil concentrations between properties within 1km of a freeway and those farther away?

```{r}
# Add a column that distinguishes concentrations within 1km of a freeway from those farther away 
soil_cores <- soil_cores |>
  mutate(dist_to_freeway_km = dist_to_freeway_m / 1000) |> 
  mutate(close = (dist_to_freeway_km <= 1))
soil_cores 

# Summarize to get the difference in mean distance
soil_mean_distance <- soil_cores |> 
    summarize(means_by_distance = mean(mean_pb_ppm), .by = close) |> 
    arrange(close)
soil_mean_distance

# Find the observed difference in means
obs_diff_mean <- soil_mean_distance$means_by_distance[2] - soil_mean_distance$means_by_distance[1]
obs_diff_mean

# ASK: DID WE SUBTRACT THESE PROPERLY? HOW TO KNOW? 

# Resample observations and recalculate the statistic
samp_dist <- replicate(
  1e4,
  {
    resampled <- soil_cores |>
      slice_sample(n = nrow(soil_cores), replace = TRUE) |>
      summarize(means_by_distance = mean(mean_pb_ppm), .by = close) |>
      arrange(close)

    resampled$means_by_distance[2] -
      resampled$means_by_distance[1]
  }
)
head(samp_dist)

# Find the confidence interval
area_95ci <- quantile(samp_dist, c(0.025, 0.975))
area_95ci

# Visualize the sampling distribution
tibble(diff_means = samp_dist) |>
  ggplot(aes(diff_means)) +
  geom_histogram(binwidth = 5) + 
  geom_vline(xintercept = obs_diff_mean, color = 'magenta') +
  geom_vline(xintercept = area_95ci, linetype = 'dashed')

```

### Interpreting our results
Based on your confidence interval, how much could the difference in lead exposure between close and far away be?

2.5%     97.5% 
 40.70573 102.95075 

Def of confidence interval - Range of values you are confident contain the pop parameter. If repeated 100 times, about 95 of those confidence intervals would contain population parameter

Our confidence interval estimates that 95% of time, the difference in lead exposure between properties within 1km of a freeway and those farther away will fall within the the interval 40.7 ppm and 102.95 ppm. 

if you repeat the same sampling 100, 95% of resulting intervals will capture the true population parameter 

### ASK ABOUT THIS



## 3. Simulate a negative result

Recall that a p-value is the probability of observing your sample _if_ the null hypothesis is true. Therefore, if you draw a sample from two groups that part of the same population, you should get a p values less than 0.05 only 5% of the time. In other words, a negative result (no actual difference) will yield a significant p-value at the 0.05 threshold in 5% of samples. This is known as the **false positive rate**.

To simulate a negative result, perform the following.

1. Define a normally-distributed population. Pick any mean and standard deviation.
2. Draw a sample of 50 observations from this population. Assign the first 25 to group A and the second 25 to group B.
3. Conduct a permutation test to investigate whether the mean of group A is greater than the mean of group B.
4. Record your p-value.
5. Repeat the simulation 1,000 times.

When you're done, verify that your p-value is less than 0.05 in approximately 5% (50) of your simulations.
